---
title: "Strings and Factors"
author: "Sarahy Martinez"
date: "2024-10-16"
output: github_document
---


If we are doing a logistic regression and we have a categorical we have to assign one as a reference group and another not, so R built in features for something that looks like character or string would factor but not anymore so we have to do it our-self so we have to use the stringr package. 

Some common things 

concatenating: joining snippets inot a long string 
shortening, subsetting, or truncating 
changing cases 
replacing one string segment with another 

when you know a general pattenr but not an exact match you need to use regular expressions 
- instead of looking for the letter "a" you might look for any string that starts with a lower-case vowel 


controlling factors is critical! 
- define reference group in models 
- ordering variables in output 
- introducing new factor levels 

common factor operations 
- converting character variable factors 
- releveling by hand, or count, or by a second variable 
- renaming levels 
- dropping unused levels 

The forcats package is the way to go for majority of your factor needs 

```{r, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Load the necessary libraries. 

```{r}
library(rvest)  #data from web 
library(p8105.datasets) # examples from this package
```




## Let's do strings and regex

```{r}
string_vec = c("my", "name", "is", "jeff")

# vectors that exist in 4 elements and everything is lowercase

str_detect(string_vec, "jeff")

#is it possible to identify the element of this string that corresponds to name. Use str_detect (presence or abscence of string)
# given a string or character observation can I detect a pattern 
# will report FALSE FALSE FALSE TRUE where it is detected last 

str_detect(string_vec, "e")
# when detecting, does the pattern appear in any collection of characters kinda like control find


str_detect(string_vec, "J")

#but if you are working with strings upper case anf lowercase will appear different, case sesitive. 

str_replace(string_vec, "jeff", "Jeff")
# name should be captialized, so can replace it with str_replace with capital 

str_replace(string_vec, "e", "E")
```


For exact matches , you can designate matches at the beginning or end of file 
```{r}

# lets see diff ways we might detect in string vector
string_vec = c(
  "i think we all rule for participating",
  "i think i have been caught",
  "i think this will be quite fun actually",
  "it will be fun, i think"
  )

#see if we can detect i think 
str_detect(string_vec, "i think")

#get four trues 

str_detect(string_vec, "^i think")

#lets take string I want to match and be more specific, string that starts with I think 

str_detect(string_vec, "i think$")

# string that ends withi think 
```

```{r}

string_vec = c(
  "Y'all remember Pres. HW Bush?",
  "I saw a green bush",
  "BBQ and Bushwalking at Mongolo Gorge",
  "BUSH -- LIVE IN CONCERT !! "
)


str_detect(string_vec,"Bush" )

str_detect(string_vec,"[Bb]ush" )
# detects both upper and lowercase bush

```



You can designate a list of characters that will count as a match 
```{r}
string_vec = c(
  "Time for a Pumpkin Spice Latte!",
  "went to the #pumpkinpatch last weekend",
  "Pumpkin Pie is obviously the best pie",
  "SMASHING PUMPKINS -- LIVE IN CONCERT!!"
  )


str_detect(string_vec, "pumpkin")

str_detect(string_vec, "Pumpkin")

str_detect(string_vec, "PUMPKIN")

str_detect(string_vec, "[Pp]umpkin")
```

You don't have to list these, instead you can provide a range of letters or numbers that count as a match

```{r}
string_vec = c(
  '7th inning stretch',
  '1st half soon to begin. Texas won the toss.',
  'she is 5 feet 4 inches tall',
  '3AM - cant sleep :('
  )

str_detect(string_vec, "[0-9][a-zA-Z]")

#what if we want to detect a number first followed immediately by a letter. Second line wont count becuase we have a number + space then letter

```

The character ( .) matches anything 
A lil weirder ...

```{r}
string_vec = c(
  'Its 7:11 in the evening',
  'want to go to 7-11?',
  'my flight is AA711',
  'NetBios: scanning ip 203.167.114.66'
  )

str_detect(string_vec, "7.11")
# we wanna detect all the 711, use the dot to separate the characters, and match anything so we dont need the carrot, except for the third because there is no special character in between 

# we want it to detect the dot like an actual dot legit , not another character


str_detect(string_vec, "7\\.1")
```

Some characters are "special". These include [ and ] , (and), and .  . If you want to search for these, you have to use the indicate they are special using \. Unfortunately, \ is also special so things start to get real strange


```{r}
string_vec = c(
  'The CI is [2, 5]',
  ':-]',
  ':-[',
  'I found the answer on pages [6-7]'
  )

str_detect(string_vec, "\\[")

#if we want to detect an open bracket and its a special character so we have to double slash
```


There are a lot of other regular expression resources; we’re really only scratching the surface. The stuff we have already will be useful and it’s probably not worth going down the regex rabbit hole. That said, it’s helpful to know what other functions exist in stringr – you should at least know they exist even if you don’t know exactly how they work right now!

## Factors ...

Factors are maybe the least intuitive of R’s data types. They are very useful, but they also do things you don’t expect; this is especially bad when you have factors but think you have characters – which happens more than you’d expect, because R uses factors a lot (mostly for historical reasons). Folks get pretty riled up about factors.

Factors are the way to store categorical variables in R. They can take on specific levels (e.g. male and female) which are usually presented as characters but are, in fact, stored by R as integers. These integer values are used by functions throughout R – in making plots, in organizing tables, in determining the “reference” category – but most of the time are hidden by easier-to-read character string labels. This close relationship to strings, when in fact there is a lot of added structure, is why factors can be so confusing.

This is the kind of thing that can get you in trouble





```{r}
sex_vec = factor(c("male", "male", "female", "female"))

#given a collection R has implemented an order, uses alphabetical 



as.numeric(sex_vec)

# takes your levels and saying underlying male is 2 and female is a 1 
```
What happens if I relevel some..
do some releveling ...

```{r}
sex_vec = fct_relevel(sex_vec, "male")

# made male the first category, male is listed as first level and female second 

as.numeric(sex_vec)

# now male is 1 and female 2
# factors and behind scenes is most confusing and something R uses , they show up as numeric elements. Exist and matter but dont really see whats going on. 


```
The previous code also illustrates coersion: forcing a variable from one type (e.g. factor) to another (e.g. numeric). Understanding how R coerces variables is important, because it sometimes happens unintentionally and can break your code or impact your analyses.

For this reason, it’s important to be deliberate factors, particularly by using functions in forcats.


## Revisit a lot of examples!!

NSDUH

We’ll revisit the table scraped from the National Survey on Drug Use and Health data on this page. In reading data from the web, we loaded this data using the code below, but noted it wasn’t tidy.

Reading data from the web

```{r}
url = "http://samhda.s3-us-gov-west-1.amazonaws.com/s3fs-public/field-uploads/2k15StateFiles/NSDUHsaeShortTermCHG2015.htm"

drug_use_html = read_html(url)

marj_use_df = 
  drug_use_html |> 
  html_table() |> 
  first() |> 
  slice(-1) 

# notice it is not a tidy dataset we the columns dont follow naming conventions etc. 

```

Get the pieces I actually need.

There are a few steps we need to implement to tidy these data. For now I’m not interested in the p-values (I’d rather just see the data); we also have age groups and year ranges in column titles, both of which are, in fact, variables. Lastly, the table includes letters as superscripts next to table entries; if we only want the percents we’ll need to strip these out.

```{r, eval =FALSE}

#tidying the data lines html to slice are reading in 
marj_use_df = 
  drug_use_html |> 
  html_table() |> 
  first() |> 
  slice(-1) |> 
  select(-contains("P Value")) |>   #removing everything that contain ( doing similar to str detect and giving exact string) 
  pivot_longer(
    cols = -State,
    names_to = "age_year",
    values_to = "percent"
  ) |> 
  separate(age_year, into = c("age", "year"), sep = "\\(") |>  # separating by open parenthesis,doesn't work bc its a special character \\
  mutate(
    year = str_replace(year, "\\)", ""), # we are updating now, we have trailing open parenthesis so we will replace 
    percent = str_remove(percent, "[a-c]$"), #updating the percent by removing the letters but be careful 
    percent = as.numeric(percent)
  ) 

# other way 
other_data_marj = 
  marj_use_df |>
  select(-contains("P Value")) |>
  pivot_longer(
    -State,
    names_to = "age_year", 
    values_to = "percent") |>
  separate(age_year, into = c("age", "year"), sep = "\\(") |>
  mutate(
    year = str_replace(year, "\\)", ""),
    percent = str_replace(percent, "[a-c]$", ""), #updating the percent by removing the letters but be careful 
    percent = as.numeric(percent)) |>
  filter(!(State %in% c("Total U.S.", "Northeast", "Midwest", "South", "West")))
```


We used stringr and regular expressions a couple of times above:

in separate, we split age and year at the open parentheses using "\\("
we stripped out the close parenthesis in mutate
to remove character superscripts, we replaced any character using "[a-c]$"
Let’s quickly visualize these data for the 12-17 age group; to make the plot readable, we’ll treat State as a factor are reorder according to the median percent value.


#NSDUH as factors 
```{r, eval=FALSE}
marj_use_df |> 
  filter(
    age == "12-17",
    !(State %in% c("Total U.S.", "South"))) |>   # state is character and not a factor so will place in alphabet order
  mutate(
    State = fct_reorder(State, percent)  # the , percent will make it the priority and put it order of prevelance
  ) |> 
  ggplot(aes(x = State, y = percent, color = year)) +   # color = year will fill in years as color ,will be the legend
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))   #axis rotate so labels are crammed 

# other way 

data_marj |>
  filter(age == "12-17") |> 
  mutate(State = fct_reorder(State, percent)) |> 
  ggplot(aes(x = State, y = percent, color = year)) + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


## NYC Restaurant Inspections


strings- restaraunt name, food, boro


```{r}
data("rest_inspec")

```

```{r}
rest_inspec |> 
  count(boro, grade) |> 
  pivot_wider(
    names_from = grade,
    values_from = n
  )

# other way 

rest_inspec |> 
  group_by(boro, grade) |> 
  summarize(n = n()) |> 
  pivot_wider(names_from = grade, values_from = n)
```

To simplify things, I’ll remove inspections with scores other than A, B, or C, and also remove the restaurants with missing boro information. I’ll also clean up boro names a bit.

```{r}
rest_inspec =
  rest_inspec |> 
  filter(
    str_detect(grade, "[A-C]"),
    !(boro == "Missing")
  )

#other method 
rest_inspec =
  rest_inspec |>
  filter(grade %in% c("A", "B", "C"), boro != "Missing") |> 
  mutate(boro = str_to_title(boro))

```

Let’s focus only on pizza places for now, and re-examine grades by borough.

```{r}
rest_inspec |> 
  mutate(dba = str_to_sentence(dba)) |> 
  filter(str_detect(dba, "Pizza"))


```


```{r}
#other example 
rest_inspec |> 
  filter(str_detect(dba, "Pizza")) |> 
  group_by(boro, grade) |> 
  summarize(n = n()) |> 
  pivot_wider(names_from = grade, values_from = n)

# That doesn’t look right – for sure there are more pizza place ratings than that! The problem is that the match in str_detect is case-sensitive until we tell it not to be:
rest_inspec |> 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) |> 
  group_by(boro, grade) |> 
  summarize(n = n()) |> 
  pivot_wider(names_from = grade, values_from = n)
```


The table is okay but I’d like to visualize this.

```{r}
rest_inspec |> 
  mutate(dba = str_to_sentence(dba)) |> 
  filter(str_detect(dba, "Pizza")) |> 
  mutate(boro = fct_infreq(boro)) |> 
  ggplot(aes(x = boro)) + 
  geom_bar()


#other example 

rest_inspec |> 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) |>
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar() 

```

Might help to have things in a different order – maybe number of pizza places? Fortunately this can be done using fct_infreq.

```{r}
rest_inspec |> 
  mutate(dba = str_to_sentence(dba)) |> 
  filter(str_detect(dba, "Pizza")) |> 
  mutate(boro = fct_relevel(boro, "STATEN ISLAND")) |> 
  ggplot(aes(x = boro)) + 
  geom_bar()

# other example 
rest_inspec |> 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) |>
  mutate(boro = fct_infreq(boro)) |>
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar() 


```

Renaming a borough 

```{r}
rest_inspec |> 
  mutate(dba = str_to_sentence(dba)) |> 
  filter(str_detect(dba, "Pizza")) |> 
  mutate(
    boro = fct_infreq(boro),
    boro = str_replace(boro, "MANHATTAN", "THE CITY")
    ) |> 
  ggplot(aes(x = boro)) + 
  geom_bar()


# other example 
rest_inspec |> 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) |>
  mutate(
    boro = fct_infreq(boro),
    boro = str_replace(boro, "Manhattan", "The City")) |>
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar() 

# if we use base R 
rest_inspec |> 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) |>
  mutate(
    boro = fct_infreq(boro),
    boro = replace(boro, which(boro == "Manhattan"), "The City")) |>
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar()

#That didn’t work at all! Factors have very specific values, trying to use a value that is not an existing factor level won’t work.

#Fortunately there is a dedicated function for renaming factor levels:

rest_inspec |> 
  filter(str_detect(dba, regex("pizza", ignore_case = TRUE))) |>
  mutate(
    boro = fct_infreq(boro),
    boro = fct_recode(boro, "The City" = "Manhattan")) |>
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar()

```
This renamed the borough, but then converted the result back to a string – which, when plotted, was implicitly made a factor and ordered alphabetically. I could switch the order in which I rename and encode the factor order I want, but that might not always work.

```{r}
rest_inspec |> 
  mutate(dba = str_to_sentence(dba)) |> 
  filter(str_detect(dba, "Pizza")) |> 
  mutate(
    boro = fct_infreq(boro),
    boro = fct_recode(boro, "THE CITY" = "MANHATTAN"),
    ) |> 
  ggplot(aes(x = boro)) + 
  geom_bar()
```


One last thing on factors...

```{r}
rest_inspec |> 
  mutate(dba = str_to_sentence(dba)) |> 
  filter(str_detect(dba, "Pizza")) |> 
  mutate(boro = fct_infreq(boro)) |> 
  lm(zipcode ~ boro, data = _)
```



# WEATHER DATA 


We saw factors in in Viz Pt 2. In that case, we were interested in reordering a factor variable in order to produce clearer plots – ggplot uses factor levels to determine the order in which categories appear. Let’s revisit those examples now.

Our first step is to load the data we used.


```{r}

weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2023-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())


```


```{r}
weather_df |>
  mutate(name = forcats::fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |> 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```


We could instead reorder name according to tmax values in each name using fct_reorder:

```{r}
weather_df |>
  mutate(name = forcats::fct_reorder(name, tmax)) |> # re ordering based on tmax
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```



# Linear Regression
Although you may not have seen linear regression previously, the ordering of factor variables play an important in this case as well. Specifically, the ordering determines the “reference” category, and is something that can be adjusted as needed.


Fit linear model where tmax is our outcome and weather station name is a preditor, should fit a linear regression 


```{r}
weather_df |>
  lm(tmax ~ name, data = _)

#r needs to know reference group so we need to factor bc in first we dont see central park bc its a reference. Now we factor to change the references.

weather_df |>
  mutate(name = forcats::fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |> 
  lm(tmax ~ name, data = _)


```

# Pulse Data 

Let’s revisit a dataset we’ve seen a few times now. In tidy data we spent some time tidying the PULSE dataset. As part of that we used tools available to us at the time, but now we have some better tools.

The code below updates the data processing pipeline for the PULSE data using stringr and forcats. The result is the same, and the differences are pretty small, but this is a bit cleaner.


```{r, eval =FALSE}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  select(id, visit, everything()) |>
  mutate(
    visit = str_replace(visit, "bl", "00m"),
    visit = factor(visit)) |>
  arrange(id, visit)

print(pulse_data, n = 12)


```


# Airbnb

We could also use factors in an exploratory analysis of the Airbnb data, for example when looking at the distribution of prices in various neighborhoods. Ordering these according to the median price makes for clearer plots than ordering neighborhoods alphabetically.

```{r, eval=FALSE}
data("nyc_airbnb")

nyc_airbnb |>
  filter(neighbourhood_group == "Manhattan") |> 
  mutate(
    neighbourhood = fct_reorder(neighbourhood, price)) |> 
  ggplot(aes(x = neighbourhood, y = price)) +
  geom_boxplot() +
  coord_flip() + 
  ylim(0, 1000)


```
















